beautier: BEAUti for R

### 2017-12-20 article.tex
beautier in verbatim
First go

\begin{abstract}
  \textbf{1. }
  Here, I present a package, \verb;beautier;, for the R programming language. \\
  \textbf{2. }
    \verb;beautier; allows for scripted use of the BEAST2 phylogenetics tool, 
    by creating BEAST2 input files from an R function call. \\
  \textbf{3. }
    I describe \verb;beautier; usage, the novel functionality it provides
    compared to BEAUti, and give some minimal examples. \\
  \textbf{4. }
    As \verb;beautier; is free, libre, open-source and designed to be extended, 
    I conclude by describing the current development of the package \\
\end{abstract}

@@

BEAST2 \cite{bouckaert2014beast} is a popular Bayesian phylogenetics tool.
BEAST2 is bundled with the program called BEAUti \cite{drummond2012bayesian},
which  allows for a user-friendly way to create the
input files that BEAST2 needs. BEAUti is convenient to get started, yet
does not allow for the scripted use of BEAST2. For creating BEAST2 input
files from a script, \verb;beautier; is the R package facilitating this:
it creates a BEAST2 input files from an R function call.

@@

The goal of the R package \verb;beautier; is to create the 
same BEAST2 input files that BEAUti creates, but from one R function call 
instead.

@@

\verb;beautier; has only parts of the functionality of BEAUti, yet
can be extended easily. beastcriptr, on the other hand, does allow for specifying
to use phylogenies of fixed crown age.

@@

As in BEAUti, to create a BEAST2 input file, 
one needs an alignment and model settings.
With those settings at their default, \verb;beautier; is easy to use:
Create a BEAST2 input file with name \verb;beast2.xml; from 
a FASTA file with name \verb;alignment.fas; is done like this:

\begin{lstlisting}[language=R, caption=Simplest example]
library(beautier)
create_beast2_input_file(
  "alignment.fas",
  "beast2.xml"
)
\end{lstlisting}

The created file \verb;beast2.xml; can then be used by BEAST2.

Novel about \verb;beautier; is that it allows for specifying a fixed crown age:

\begin{lstlisting}[language=R, caption=Example with fixed crown age]
create_beast2_input_file(
  "alignment.fas",
  "beast2.xml"
  fixed_crown_age = TRUE,
  initial_phylogeny = beautier::fasta_to_phylo(
    fasta_filename = "alignment.fas", 
    crown_age = 15
  )
)
\end{lstlisting}

In this example, a random phylogeny is created from the FASTA file, but
one could also use a more informed initial phylogeny with the desired crown age.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{\verb;beautier; development and other resources}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\verb;beautier; successfully creates valid BEAST2 input files. Unlike BEAUti,
beascriptr allows for setting a fixed crown age.

@@

\verb;beautier; has minimal support for calling BEAST2 from within R and does
so for testing purposes.

>>>>

\begin{abstract}
  \textbf{1. }
  Here, I present a package, \verb;beautier;, 'BEAUti for R', for the R programming language. \\
  \textbf{2. }
    \verb;beautier; allows for scripted use of the BEAST2 phylogenetics tool, 
    by creating BEAST2 input files from an R function call. \\
  \textbf{3. }
    I describe \verb;beautier; usage, the novel functionality it provides
    compared to BEAUti, and give some minimal examples. \\
  \textbf{4. }
    As \verb;beautier; is free, libre, open-source and designed to be extended, 
    I conclude by describing the current development of the package \\
\end{abstract}

@@

BEAST2 \cite{bouckaert2014beast} is a popular Bayesian phylogenetics tool.
BEAST2 is bundled with the program called BEAUti \cite{drummond2012bayesian}.
BEAUti has a user-friendly graphical user interface to create BEAST2 input files,
but does not allow for the scripted use of BEAST2. '
\verb;beautier; does allow for scripted use of BEAST2 from an R function call.
An additional novelty is that \verb;beautier; allows to specify a fixed crown age.

@@

\verb;beautier; is written in the R programming language \cite{R}.
\verb;beautier; creates the BEAST2 input files from an R function call.

\verb;beautier;'s main function is \verb;create_beast2_input_file;, which creates
an BEAST2 input file.

\verb;create_beast2_input_file; needs at least the name of a FASTA file containing a DNA alignment
and a name for the to-be-created output file. This interface follows BEAUti's default settings.
Per alignment, a site model, clock model and tree priors can be chosen.
Multiple alignments can be used, each with its own (unlinked) site model, clock model and tree prior.

\begin{table}[]
\centering
\begin{tabular}{ | l | l | }
\hline
\textbf{Name} & \textbf{Description} \\
\hline
\verb;create_beast2_input_file; & Creates a BEAST2 input file \\
\hline
\verb;create_gamma_site_model; & Create a gamma site model \\
\verb;create_gtr_site_model; & Create a GTR site model \\
\verb;create_hky_site_model; & Create an HKY site model \\
\verb;create_jc69_site_model; & Create a Jukes-Cantor site model \\
\verb;create_tn93_site_model; & Create a TN93 site model \\
\hline
\verb;create_rln_clock_model; & Create a relaxed log-normal clock model \\
\verb;create_strict_clock_model; & Create a strict clock model \\
\hline
\verb;create_bd_tree_prior; & Create a birth-death tree prior \\
\verb;create_cbs_tree_prior; & Create a coalescent Bayesian skyline tree prior \\
\verb;create_ccp_tree_prior; & Create a coalescent constant-population tree prior \\
\verb;create_cep_tree_prior; & Create a coalescent exponential-population tree prior \\
\verb;create_yule_tree_prior; & Create a Yule tree prior \\
\hline
\verb;create_beta_distr; & Create a beta distribution \\
\verb;create_exp_distr; & Create an exponential distribution \\
\verb;create_gamma_distr; & Create a gamma distribution \\
\verb;create_inv_gamma_distr; & Create an inverse gamma distribution \\
\verb;create_laplace_distr; & Create a Laplace distribution \\
\verb;create_log_normal_distr; & Create a log-normal distribution \\
\verb;create_normal_distr; & Create a normal distribution \\
\verb;create_one_div_x_distr; & Create a 1/X distribution \\
\verb;create_poisson_distr; & Create a Poisson distribution \\
\verb;create_uniform_distr; & Create a uniform distribution \\
\hline
\end{tabular}
\caption{beautier's functions}
\label{tab:functions}
\end{table}

@@

\verb;beautier; has only parts of the functionality of BEAUti, yet
can be extended easily. \verb;beautier; , on the other hand, does allow for specifying
to use phylogenies of fixed crown age.

@@

In R, a package's function need to be loaded in the global namespace first:

\begin{lstlisting}[language=R, caption=Loading, label=lst:loading_beautier, floatplacement=H]
library(beautier)
\end{lstlisting}

BEAUti, and likewise \verb;beautier;, need at least a FASTA filename
and an XML output filename. In BEAUti, this is achieved by loading a FASTA file (resulting
in figure \ref{fig:simplest_beauti_usage}), then saving an output file using a common
save file dialog. In \verb;beautier;, the same is achieved by listing \ref{lst:simplest_example}:

\begin{lstlisting}[language=R, caption=Simplest example, label=lst:simplest_example, floatplacement=H]
library(beautier)
create_beast2_input_file(
  "alignment.fas",
  "beast2.xml"
)
\end{lstlisting}

\begin{figure}
  \centering
  \includegraphics[]{all_default.png}
  \caption{Simplest BEAUti usage}
  \label{fig:simplest_beauti_usage}
\end{figure}

This code will create a BEAST2 file with name '\verb;beast2.xml;',
using a FASTA file with name \verb;alignment.fas;, using the same default settings as BEAUti.
The default settings are, among others, to use a Jukes-Cantor site model, a strict clock,
and a Yule birth tree prior. Listing \ref{lis:simplest_example_explicit} shows how
to explicitly pick these settings:

\begin{lstlisting}[language=R, caption=Simplest example with explicit defaults, label=lst:simplest_example_explicit, floatplacement=H]
library(beautier)
create_beast2_input_file(
  input_fasta_filenames = "alignment.fas",
  output_xml_filename = "beast2.xml",
  site_models = create_jc69_site_model(),
  clock_models = create_strict_clock_model(),
  tree_priors = create_yule_tree_prior()
)
\end{lstlisting}

The argument names \verb;input_fasta_filenames;, \verb;site_models;, \verb;clock_models; and \verb;tree_priors; are plural, as each of these
can be (a list of) one or more elements. Each of these arguments must have the same number of elements, so that each alignment has its
own site model, clock model and tree prior. An example of using a different site model, clock model and tree prior is shown by listing \ref{lis:all_different}:

\begin{lstlisting}[language=R, caption=Example with different site model and clock model and tree prior, label=lst:all_different, floatplacement=H]
library(beautier)
create_beast2_input_file(
  input_fasta_filenames = "alignment.fas",
  output_xml_filename = "beast2.xml",
  site_models = create_hky_site_model(),
  clock_models = create_rln_clock_model(),
  tree_priors = create_bd_tree_prior()
)
\end{lstlisting}

This code uses an HKY site model, a relaxed log-normal clock model and a birth-death tree prior.
Table \ref{tab:functions} shows an overview of all functions to create site models, clock models and tree priors.

\verb;beautier; creates site models, clock models and tree priors with the same default distrbutions as BEAUti.
For example, a Yule tree prior assumes that birth rate likelihoods follow a uniform distribution, from minus infinity
to infinity. This assumption entails that negative and positive birth rates are just as likely, where a negative birth rate is biologically impossible. 
One may prefer to have an exponential distribution instead, as this would state that birth rates are always positive, and
higher values are less likely than lower values. To do so \verb;beautier; is shown by listing \ref{lis:diff_distr}:

\begin{lstlisting}[language=R, caption=Example with Yule tree prior with different birth rate distribution, label=lst:diff_distr, floatplacement=H]
library(beautier)
create_beast2_input_file(
  input_fasta_filenames = "alignment.fas",
  output_xml_filename = "beast2.xml",
  tree_priors = create_yule_tree_prior(
    birth_rate_distr = create_exp_distr()    
  )
)
\end{lstlisting}

Novel about \verb;beautier; is that it allows for specifying a fixed crown age. 
By default, a phylogeny's crown age is jointly estimated with the other parameters.
Setting a fixed crown age is not yet possible in BEAUti directly, but it is documented how to 
manually edit the XML file to allow for a fixed crown age. Listing \ref{lst:fixed_crown_age} shows how to specify a fixed crown age is \verb;beautier;:

\begin{lstlisting}[language=R, caption=Example with fixed crown age, label=lst:fixed_crown_age, floatplacement=H]
create_beast2_input_file(
  "alignment.fas",
  "beast2.xml"
  fixed_crown_age = TRUE,
  initial_phylogeny = fasta_to_phylo(
    fasta_filename = "alignment.fas", 
    crown_age = 15
  )
)
\end{lstlisting}

This code shows that setting \verb;fixed_crown_age; to true is insufficient. An
initial phylogeny of the desired (fixed) crown age needs to be supplied. In this
example, a random phylogeny is constructed from the FASTA filename. Supplying an 
informed phylogeny will make the MCMC algorithm start at a likelier initial state.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{beautier development and other resources}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\verb;beautier; successfully creates valid BEAST2 input files. Unlike BEAUti,
\verb;beautier; allows for setting a fixed crown age.

@@

\verb;beautier; has minimal support for calling BEAST2 from within R and does
so for testing purposes only.
